---
title: R Coding Basics
#subtitle: 
date: today
date-format: long
footer: "[DATS 1001 Website](https://ds4all.rocks)"
logo: images/ds4all-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

# R Programming Basics

## What Can R Do? 

<br>

::: incremental
- R is a powerful language for data analysis and visualization
- It is also a general-purpose programming language
- Does everything from web development to machine learning
- It is open-source and has a large community of users and developers
:::

## R as a Calculator

<br>

- R can be used as a simple calculator
- You can perform arithmetic operations on numbers

```{r}
#| label: arithmetic
#| echo: true

# Addi a number and store it to a value
sum_of_2plus2 <- 2 + 2


sum_of_2plus2
```

## Some Common Arithmetic Operators

<br>

- `+` addition
- `-` subtraction
- `*` multiplication
- `/` division
- `^` exponentiation (also `**`)

## Functions

- A function is a set of instructions that produces some output
- In R, you can use built-in functions to perform specific tasks
- For example, you can use the `mean()` function to calculate the average of a set of numbers
- To do this you have to use the combine function `c()` to create a vector of numbers

## 

<br> 

Create a vector of numbers and take the mean...

<br>

```{r}
#| label: mean
#| echo: true

# Create a vector of numbers
numbers <- c(1, 2, 3, 4, 5)

# Calculate the mean of the numbers
mean(numbers)
```

## Some Common Base R Functions {.smaller}

- `mean()` calculates the mean of a set of numbers
- `median()` calculates the median of a set of numbers
- `sd()` calculates the standard deviation of a set of numbers
- `sum()` calculates the sum of a set of numbers
- `length()` calculates the length of a vector
- `max()` and `min()` calculate the maximum and minimum values of a vector
- `round()` rounds a number to a specified number of decimal places
- `sqrt()` calculates the square root of a number
- `log()` calculates the natural logarithm of a number
- `exp()` calculates the exponential of a number
- `abs()` calculates the absolute value of a number

# Packages

## From Functions to Packages {.smaller}

<br>

::: {.incremental}
-   A *function* is a set of instructions
    -   `read_csv()` is a function
    -   `ggplot()` is a function
-   A *package* is a collection of functions
    -   `readr` is a package that contains the `read_csv()` function
    -   `ggplot2` is a package that contains the `ggplot()` function
-   Use `install.packages()` to install packages
-   Use `library()` to load packages
-   You can install packages from CRAN
:::

## The Tidyverse {.smaller}

::: {.incremental}
-   The Tidyverse is a collection of data science packages
-   It is also considered a dialect of R
-   In this class, we will be using many Tidyverse packages
    -   `ggplot2` for data visualization
    -   `readr` for reading data
    -   `dplyr` for data manipulation
    -   `tidyr` for data tidying
    -   Etc.
-   At first we will load the packages independently, e.g. `library(ggplot2)`
-   Later we will load them all at once with `library(tidyverse)`
:::

# Objects

## What is an Object?

::: incremental
- An object in R is a data structure used to store data
- It can vary from simple scalar types to more complex data structures like vectors, lists, or data frames
- Objects hold not only data but also information about the type of data and the operations that can be performed on them
- Every entity in R is considered an object, making R a language based around the manipulation of objects
:::

## How to Store Data

- In R, you can store data in objects using the assignment operator `<-`
- The object name is on the left of `<-`, and the data or value you wish to assign to the object is on the right
- Then you can print the object to the console using the object name

```{r}
#| label: objects
#| echo: true

# Store the value 42 in the object my_number
my_number <- 42

# Print the value of my_number
my_number 
```

## When to Store Data in Objects

<br>

- Note that you don't always have to store data in objects
- You should mostly store data in objects when you want to use the data later
- If you only need to use the data once, you can just use the data directly

```{r}
#| label: arithmetic2
#| echo: true

# Add two numbers without storing them in an object
2 + 2
```

## Your Turn!

<br>

- Go to the Week 2 module on Posit Cloud
- Work through the first set of challenges there

```{r}
#| label: timer0

library(countdown)
countdown(minutes = 5, 
          id = "timer0", 
          bottom = "10%",
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

# Three Examples

## Let's get going . . .

<br>

Let's open up the Getting Started module on Posit Cloud...

And work through the examples there.

## Example: Make a map!

```{r}
#| label: leaflet_map

library(leaflet)
leaflet() %>% 
  addTiles() %>%   # Add default OpenStreetMap map tiles
  addMarkers(lat = 38.90243843683386, lng =  -77.0443814477152, 
             label = "Elliott School of International Affairs")
```

## Example: Make a map!

<br>

```{r}
#| label: leaflet_map2
#| echo: true
#| eval: false
  
library(leaflet)
leaflet() %>% 
  addTiles() %>%   # Add default OpenStreetMap map tiles
  addMarkers(lat = 38.90243843683386, lng =  -77.0443814477152, 
             label = "Elliott School of International Affairs")
```

```{r}
#| label: timer1

library(countdown)
countdown(minutes = 5, 
          id = "timer1")
```

## Example: Plotting Democracy Over Time

```{r}
#| label: plot_democracy

# Load the packages
library(vdemlite)
library(ggplot2)

# Use vdemlite to extract democracy scores for France and INdia
dem_data <- fetchdem(indicators = "v2x_polyarchy",
                     countries = c("FRA", "IND"))

# And now we can plot the data
ggplot(dem_data, aes(y = v2x_polyarchy, x = year, color=country_name)) +
  geom_line() +
  theme_minimal() +
  xlab("Year") +
  ylab("Electoral Democracy Index") +
  ggtitle("Electoral Democracy, 1970-2022") +
  geom_hline(yintercept = .5, linetype = "dashed", color = "grey") +
   scale_color_manual(name="Country", values=c("#E69F00", "#56B4E9")) +
  ylim(0, 1)
```

## Example: Plotting Democracy Over Time

```{r}
#| label: plot_democracy2
#| echo: true
#| eval: false

# Load the packages
library(vdemlite)
library(ggplot2)

# Use vdemlite to extract democracy scores for France and INdia
dem_data <- fetchdem(indicators = "v2x_polyarchy",
                     countries = c("FRA", "IND"))

# And now we can plot the data
ggplot(dem_data, aes(y = v2x_polyarchy, x = year, color=country_name)) +
  geom_line() +
  theme_minimal() +
  xlab("Year") +
  ylab("Electoral Democracy Index") +
  ggtitle("Electoral Democracy, 1970-2022") +
  geom_hline(yintercept = .5, linetype = "dashed", color = "grey") +
   scale_color_manual(name="Country", values=c("#E69F00", "#56B4E9")) +
  ylim(0, 1)
```

```{r}
#| label: timer2

library(countdown)
countdown(minutes = 5, 
          id = "timer2",
          bottom = "0%",
          right = "10%")
```

## Example: UN Voting Trends

```{r}
#| label: un_voting

library(tidyverse)
library(unvotes)
library(lubridate)
library(scales)
library(pacman)

unvotes <- un_votes %>%
  inner_join(un_roll_calls, by = "rcid") %>%
  inner_join(un_roll_call_issues, by = "rcid")

unvotes %>%
  # then filter out to only include the countries we want
  filter(country %in% c("South Africa", "United States", "France")) %>%
  # then make sure R understands the the year variable is a data
  mutate(year = year(date)) %>%
  # Then group the data by country and year
  group_by(country, year, issue) %>%
  # then take the average Yes votes for each country_year
  summarize(percent_yes = mean(vote == "yes")) %>%
  # then make a nice plot
  ggplot(mapping = aes(x = year, y = percent_yes, color = country)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~issue) +
  scale_y_continuous(labels = percent) +
  scale_color_manual( values=c("#E69F00", "#56B4E9", "#009E73")) +
  labs(
    title = "Percentage of 'Yes' votes in the UN General Assembly",
    subtitle = "1946 to 2019",
    y = "% Yes",
    x = "Year",
    color = "Country"
  )
```

## Example: UN Voting Trends

```{r}
#| label: un_voting2
#| echo: true
#| eval: false

library(tidyverse)
library(unvotes)
library(lubridate)
library(scales)
library(pacman)

unvotes <- un_votes %>%
  inner_join(un_roll_calls, by = "rcid") %>%
  inner_join(un_roll_call_issues, by = "rcid")

unvotes %>%
  # then filter out to only include the countries we want
  filter(country %in% c("South Africa", "United States", "France")) %>%
  # then make sure R understands the the year variable is a data
  mutate(year = year(date)) %>%
  # Then group the data by country and year
  group_by(country, year, issue) %>%
  # then take the average Yes votes for each country_year
  summarize(percent_yes = mean(vote == "yes")) %>%
  # then make a nice plot
  ggplot(mapping = aes(x = year, y = percent_yes, color = country)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~issue) +
  scale_y_continuous(labels = percent) +
  scale_color_manual( values=c("#E69F00", "#56B4E9", "#009E73")) +
  labs(
    title = "Percentage of 'Yes' votes in the UN General Assembly",
    subtitle = "1946 to 2019",
    y = "% Yes",
    x = "Year",
    color = "Country"
  )
```

```{r}
#| label: timer3

library(countdown)
countdown(minutes = 5, 
          id = "timer3",
          bottom = "8",
          right = "10%")
```
